---
title: Préparation des données utilisées pour le rapport
format: html
---

```{r}

library(purrr)

```

## Données "bénéficiaires ASE"

### Acquisition des données

```{r}
#| eval: false

urls_each_year <- list(
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2010_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2011_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2012_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2013_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2014_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2015_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2016_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2017_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2018_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2019_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2020_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2021_xlsx/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_en_2022_xlsx/"
)

url_time_series <- "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/les-beneficiaires-de-l-aide-sociale-a-l-enfance/attachments/les_beneficiaires_de_l_aide_sociale_a_l_enfance_series_longues_2002_2022_xlsx/"

safe_download <- safely(\(x, y) download.file(x, y, mode = "wb"))

# Creating names for the files to supply to the download.file function
fnames <- paste0("dataset-year-", 2010:2022, ".xlsx")

# Download all excel files for individual years in corresponding data/raw
# directory
map2(urls_each_year,
     paste0(file.path("..", "data", "raw", "01-dataset-beneficiaires-ase", ""),
            fnames),
     safe_download)

# Download excel file for time series between 2002 and 2022
fname <- paste0(file.path("..", "data", "raw", "01-dataset-beneficiaires-ase", ""),
         "dataset-time-series-2002-2022.xlsx")
safe_download(url_time_series, fname)

```

### Préparation des données

D'abord j'enlève les feuillets inutiles et je reformatte les tables pour ne
garder que les lignes qui m'intéressent pour tous les fichiers `.xlsx`.
Etape faite avec LibreOffice en enregistrant les feuillets
utiles en CSV (*NB:* cocher "Edit filter options" au moment de la sauvegarde).

Une fois cette étape réalisée, j'obtiens:

```
# yvelines_data/
.
├── dataset-time-series-2002-2022-total-enfants-accueillis.csv
├── dataset-time-series-2002-2022-total-enfants-confies.csv
├── dataset-year-2010.csv
├── dataset-year-2011.csv
├── dataset-year-2012.csv
├── dataset-year-2013.csv
├── dataset-year-2014.csv
├── dataset-year-2015.csv
├── dataset-year-2016.csv
├── dataset-year-2017.csv
├── dataset-year-2018.csv
├── dataset-year-2019.csv
├── dataset-year-2020.csv
├── dataset-year-2021.csv
└── dataset-year-2022.csv
```

Maintenant je peux utiliser ces fichiers CSV pour construire une série
temporelle avec le total des enfants confiés à l'ASE de 2002-2009 (avant 2002 les données départementales ne sont pas données) + la
répartition par tranche d'âge de 2010-2022. A noter que pour la répartition par
tranche d'âge, à partir de 2015 on a les catégories "<3" et "3-5" en plus des
catégories à la place de "<6" et en plus des catégories "6-10", "11-15",
"16-17", "18+" disponibles pour toutes les périodes.

## Données brutes enquête ASE

### Acquisition des données

```{r}
#| eval: false

urls_each_year <- list(
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/bases-de-donnees-brutes-de-l-enquete-aide-sociale-volet-aide-sociale-a-l-enfance/attachments/ase_brut_2018_zip/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/bases-de-donnees-brutes-de-l-enquete-aide-sociale-volet-aide-sociale-a-l-enfance/attachments/ase_brut_2019_zip/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/bases-de-donnees-brutes-de-l-enquete-aide-sociale-volet-aide-sociale-a-l-enfance/attachments/ase_brut_2020_zip/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/bases-de-donnees-brutes-de-l-enquete-aide-sociale-volet-aide-sociale-a-l-enfance/attachments/ase_brut_2021_zip/",
  "https://data.drees.solidarites-sante.gouv.fr/api/datasets/1.0/bases-de-donnees-brutes-de-l-enquete-aide-sociale-volet-aide-sociale-a-l-enfance/attachments/ase_brut_2022_zip/"
)

safe_download <- safely(\(x, y) download.file(x, y, mode = "wb"))

# Creating names for the files to supply to the download.file function
fnames <- paste0("dataset-year-", 2018:2022, ".zip")

# Download all excel files for individual years in corresponding data/raw
# directory
map2(urls_each_year,
     paste0(file.path("..", "data", "raw", "02-dataset-donnes-brutes-ase", ""),
            fnames),
     safe_download)

```

### Préparation des données

```bash

unzip "dataset-year-20*"

mv Dictionnaire-ASE-20* ../../metadata/dataset-donnes-brutes-ase

```

First, I filter the lines with the data for the Yvelines and I save them in new
CSV in `yvelines_data/`.
Now, I need to identify the columns with the data I am interested in using the
corresponding "Dictionnary" in the `metadata/` folder.

Visidata tricks I used to clean up the files:

1. Create "transposed" sheet from the original CSV: `shift-t`.
2. Open columns sheet: `shift-c`.
3. Select desired columns for girls with `g|<type:_G>` and for boys with
   `g|<type:_H>`.
4. Invert selection with `gt` and delete all selected rows with `gd`.
5. Add new columns for year and gender: `a` to append new blank column
   (`x` and `shift-p` to put column above another one) and `q` to return to
   transposed sheet
6. Select all rows `gs` and in `year` column: `ge` to put proper year value.
   Unselect everything with `gu`.
7. On `age_category` column, `g|` and type `_G`, then move to `gender` to fill
   gender value with `ge<type:F>`. For boys: `gt` to invert selection and
   `ge<type:M>`.
8. Save with `ctrl-s`. Then, quit everything and reopen new file.
9. To copy age category values: `gs` on an already formatted column from a
   previous year, `gzy` to copy cells and on destination file `gs` and `gzp` to
   overwrite cells.
